# (PART) MICROBIOME STATS {-}

# How do I statistically compare OTU richness between groups?

## Explanation

Once you’ve calculated **alpha diversity** (e.g., observed OTUs per sample), it’s common to test whether **groups differ significantly**. This can help you determine if an experimental condition affects microbial richness.

Typical tests include:
- **T-test** (for two groups, normally distributed data)
- **Wilcoxon rank-sum test** (non-parametric)
- **ANOVA or Kruskal-Wallis** (for 3+ groups)

Here we demonstrate group comparison for **richness** using appropriate statistical tests.

## Python Code
```python
import pandas as pd
from scipy.stats import ttest_ind, mannwhitneyu

# Load richness and metadata
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0)
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")

# Compute richness
richness = pd.DataFrame({
    "sample_id": otu_df.columns,
    "richness": (otu_df > 0).sum(axis=0).values
})
data = pd.merge(richness, meta_df, on="sample_id")

# Split by group
control = data[data["group"] == "Control"]["richness"]
treatment = data[data["group"] == "Treatment"]["richness"]

# T-test (assumes normality)
t_stat, t_pval = ttest_ind(control, treatment)

# Wilcoxon (non-parametric)
w_stat, w_pval = mannwhitneyu(control, treatment)

print(f"T-test p-value: {t_pval:.4f}")
print(f"Wilcoxon test p-value: {w_pval:.4f}")
```

## R Code
```{r}
library(tidyverse)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")

# Calculate richness
richness <- colSums(otu_df > 0)
richness_df <- data.frame(sample_id = names(richness), richness = richness)
merged <- left_join(richness_df, meta_df, by = "sample_id")

# T-test
t_test <- t.test(richness ~ group, data = merged)

# Wilcoxon test
wilcox_test <- wilcox.test(richness ~ group, data = merged)

t_test$p.value
wilcox_test$p.value
```

# How do I test for correlation between alpha diversity and age?

## Explanation

In many studies, you may want to examine whether microbial diversity is associated with continuous metadata like **age**, **BMI**, or **pH**.

Correlation tests help assess linear or monotonic relationships between variables:
- **Pearson correlation**: for linear relationships (assumes normality)
- **Spearman correlation**: for monotonic (rank-based) associations (non-parametric)

This Q&A demonstrates testing correlation between **richness** and **age**.

## Python Code
```python
import pandas as pd
from scipy.stats import pearsonr, spearmanr

# Load OTU and metadata
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0)
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")

# Compute richness
richness = pd.DataFrame({
    "sample_id": otu_df.columns,
    "richness": (otu_df > 0).sum(axis=0).values
})
data = pd.merge(richness, meta_df, on="sample_id")

# Pearson correlation
pearson_corr, pearson_pval = pearsonr(data["richness"], data["age"])

# Spearman correlation
spearman_corr, spearman_pval = spearmanr(data["richness"], data["age"])

print(f"Pearson r: {pearson_corr:.3f}, p = {pearson_pval:.4f}")
print(f"Spearman rho: {spearman_corr:.3f}, p = {spearman_pval:.4f}")
```

## R Code
```{r}
library(tidyverse)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")

# Compute richness
richness <- colSums(otu_df > 0)
richness_df <- data.frame(sample_id = names(richness), richness = richness)
merged <- left_join(richness_df, meta_df, by = "sample_id")

# Pearson correlation
cor.test(merged$richness, merged$age, method = "pearson")

# Spearman correlation
cor.test(merged$richness, merged$age, method = "spearman")
```

# How do I compare alpha diversity across 3 or more groups?

## Explanation

When comparing microbial richness across more than two groups, you can use:
- **ANOVA**: if data are normally distributed
- **Kruskal-Wallis test**: non-parametric alternative

This Q&A tests whether OTU richness differs by **body location** (e.g., gut, skin).

## Python Code
```python
import pandas as pd
from scipy.stats import f_oneway, kruskal

# Load data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0)
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")

# Compute richness
richness = pd.DataFrame({
    "sample_id": otu_df.columns,
    "richness": (otu_df > 0).sum(axis=0).values
})
data = pd.merge(richness, meta_df, on="sample_id")

# Split richness by location
groups = [group["richness"].values for name, group in data.groupby("location")]

# ANOVA
f_stat, f_pval = f_oneway(*groups)

# Kruskal-Wallis
kw_stat, kw_pval = kruskal(*groups)

print(f"ANOVA p-value: {f_pval:.4f}")
print(f"Kruskal-Wallis p-value: {kw_pval:.4f}")
```

## R Code
```{r}
library(tidyverse)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")

# Compute richness
richness <- colSums(otu_df > 0)
richness_df <- data.frame(sample_id = names(richness), richness = richness)
merged <- left_join(richness_df, meta_df, by = "sample_id")

# ANOVA
anova_res <- aov(richness ~ location, data = merged)

# Kruskal-Wallis
kruskal_res <- kruskal.test(richness ~ location, data = merged)

summary(anova_res)
kruskal_res
```

# How do I test for differences in community composition using PERMANOVA?

## Explanation

**PERMANOVA** (Permutational Multivariate Analysis of Variance) is used to test whether **beta diversity** significantly differs between groups.

It operates on a dissimilarity matrix (e.g., Bray-Curtis) and partitions variation based on experimental factors like **treatment group** or **location**.

This Q&A applies PERMANOVA to Bray-Curtis distances computed from OTU abundances.

## Python Code
```python
import pandas as pd
from skbio.diversity import beta_diversity
from skbio.stats.distance import permanova
from skbio import DistanceMatrix

# Load data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")

# Compute Bray-Curtis distance matrix
bray_dm = beta_diversity("braycurtis", otu_df.values, ids=otu_df.index)

# Format metadata
meta_df = meta_df.set_index("sample_id").loc[otu_df.index]

# Run PERMANOVA
result = permanova(distance_matrix=bray_dm, grouping=meta_df["group"], permutations=999)
print(result)
```

## R Code
```{r}
library(vegan)
library(tidyverse)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")

# Bray-Curtis distance
otu_t <- t(otu_df)
bray <- vegdist(otu_t, method = "bray")

# Match metadata
meta_df <- meta_df %>% filter(sample_id %in% rownames(otu_t)) %>% column_to_rownames("sample_id")

# Run PERMANOVA (adonis)
adonis_res <- adonis2(bray ~ group, data = meta_df, permutations = 999)

adonis_res
```

# How do I test for differential abundance of OTUs across groups?

## Explanation

Differential abundance analysis identifies OTUs that significantly differ between groups (e.g., Control vs Treatment).

While tools like **DESeq2** are used for RNA-seq and microbiome count data, simpler methods like:
- **Wilcoxon tests**
- **t-tests**
- **ANCOM / ALDEx2** (specialized tools)  
can also be used with filtered OTU data.

Here we demonstrate how to test one OTU at a time between groups.

## Python Code
```python
import pandas as pd
from scipy.stats import mannwhitneyu

# Load OTU table and metadata
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0)
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t").set_index("sample_id")
otu_df = otu_df[meta_df.index]  # ensure matching

# Perform Wilcoxon test for each OTU
results = []
for otu in otu_df.index:
    control = otu_df.loc[otu, meta_df["group"] == "Control"]
    treatment = otu_df.loc[otu, meta_df["group"] == "Treatment"]
    stat, pval = mannwhitneyu(control, treatment)
    results.append((otu, pval))

# Convert to DataFrame
df_results = pd.DataFrame(results, columns=["OTU", "p_value"])
df_results["adjusted_p"] = df_results["p_value"] * len(df_results)  # Bonferroni
df_results.sort_values("p_value").head()
```

## R Code
```{r}
library(tidyverse)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")

# Ensure sample order matches
otu_df <- otu_df[, meta_df$sample_id]

# Run Wilcoxon test for each OTU
results <- apply(otu_df, 1, function(x) {
  group <- meta_df$group
  test <- wilcox.test(x[group == "Control"], x[group == "Treatment"])
  return(test$p.value)
})

df_results <- data.frame(OTU = rownames(otu_df), p_value = results)
df_results$adjusted_p <- p.adjust(df_results$p_value, method = "bonferroni")
head(df_results[order(df_results$p_value), ])
```
