# (PART) MICROBIOME MACHINE LEARNING {-}

# How do I prepare microbiome data for machine learning?

## Explanation

Before applying machine learning, you must structure your OTU table and metadata into a form suitable for modeling.

Typical steps include:
- **Filtering**: Keep relevant OTUs/features
- **Merging**: Align OTU table with sample metadata
- **Encoding**: Set up group labels (e.g., Control = 0, Treatment = 1)
- **Splitting**: Train-test split to evaluate generalizability

This Q&A sets up data for classification.

## Python Code
```python
import pandas as pd
from sklearn.model_selection import train_test_split

# Load data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")

# Merge OTU table with metadata by sample
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")

# Define features (X) and labels (y)
X = data[otu_df.columns]  # OTU features
y = data["group"].map({"Control": 0, "Treatment": 1})  # binary encoding

# Split into training/testing sets
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Check shape
print("Train shape:", X_train.shape)
print("Test shape:", X_test.shape)
```

## R Note
```{r}
# Most ML pipelines in microbiome analysis are performed in Python.
# In R, similar workflows can be built using caret, tidymodels, or mlr3.
```

# How do I train and evaluate a Random Forest classifier on microbiome data?

## Explanation

The **Random Forest** algorithm is a popular and robust model for microbiome classification tasks due to its:
- Built-in feature importance
- Resistance to overfitting
- Non-linear modeling capability

This Q&A demonstrates how to:
- Train a Random Forest classifier
- Evaluate it using accuracy and confusion matrix
- Inspect important OTUs

## Python Code
```python
import pandas as pd
from sklearn.ensemble import RandomForestClassifier
from sklearn.metrics import accuracy_score, confusion_matrix
from sklearn.model_selection import train_test_split

# Load and prepare data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")
X = data[otu_df.columns]
y = data["group"].map({"Control": 0, "Treatment": 1})
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Train Random Forest
clf = RandomForestClassifier(n_estimators=100, random_state=42)
clf.fit(X_train, y_train)

# Predict and evaluate
y_pred = clf.predict(X_test)
acc = accuracy_score(y_test, y_pred)
cm = confusion_matrix(y_test, y_pred)

print(f"Accuracy: {acc:.2f}")
print("Confusion Matrix:")
print(cm)

# Top 5 important OTUs
feat_imp = pd.Series(clf.feature_importances_, index=X.columns)
print("Top OTUs:
", feat_imp.sort_values(ascending=False).head())
```

## R Code (caret)
```{r}
library(tidyverse)
library(caret)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")
otu_df <- otu_df[, meta_df$sample_id]
otu_df <- t(otu_df)
data <- cbind(as.data.frame(otu_df), group = meta_df$group)

# Encode group and split
data$group <- as.factor(data$group)
set.seed(42)
trainIndex <- createDataPartition(data$group, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]

# Train Random Forest
rf_model <- train(group ~ ., data = train, method = "rf", trControl = trainControl(method = "cv", number = 5))

# Predict and evaluate
pred <- predict(rf_model, newdata = test)
confusionMatrix(pred, test$group)
```

# How do I build a Logistic Regression model for microbiome classification?

## Explanation

**Logistic Regression** is a foundational classification model useful for:
- Binary prediction (e.g., Control vs Treatment)
- Interpreting OTU effects via coefficients
- Establishing baselines before more complex models

This Q&A shows how to train and evaluate a Logistic Regression model.

## Python Code
```python
import pandas as pd
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import accuracy_score, confusion_matrix
from sklearn.model_selection import train_test_split

# Load and prepare data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")
X = data[otu_df.columns]
y = data["group"].map({"Control": 0, "Treatment": 1})
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Train logistic regression model
model = LogisticRegression(max_iter=1000, random_state=42)
model.fit(X_train, y_train)

# Predict and evaluate
y_pred = model.predict(X_test)
acc = accuracy_score(y_test, y_pred)
cm = confusion_matrix(y_test, y_pred)

print(f"Accuracy: {acc:.2f}")
print("Confusion Matrix:")
print(cm)

# Inspect top OTUs by absolute coefficient magnitude
coef = pd.Series(model.coef_[0], index=X.columns)
print("Top OTUs:
", coef.abs().sort_values(ascending=False).head())
```

## R Code (caret)
```{r}
library(tidyverse)
library(caret)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")
otu_df <- otu_df[, meta_df$sample_id]
otu_df <- t(otu_df)
data <- cbind(as.data.frame(otu_df), group = meta_df$group)

# Encode group and split
data$group <- as.factor(data$group)
set.seed(42)
trainIndex <- createDataPartition(data$group, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]

# Train logistic regression
lr_model <- train(group ~ ., data = train, method = "glm", family = "binomial", trControl = trainControl(method = "cv", number = 5))

# Predict and evaluate
pred <- predict(lr_model, newdata = test)
confusionMatrix(pred, test$group)
```

# How do I train a Support Vector Machine (SVM) for microbiome classification?

## Explanation

**Support Vector Machines (SVM)** are powerful classifiers for high-dimensional biological data. They find a decision boundary (hyperplane) that maximizes class separation.

SVMs are well-suited for microbiome data because:
- They can handle many features (OTUs)
- Work with kernel functions for non-linear separation
- Often perform well with sparse data

## Python Code
```python
import pandas as pd
from sklearn.svm import SVC
from sklearn.metrics import accuracy_score, confusion_matrix
from sklearn.model_selection import train_test_split

# Load and prepare data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")
X = data[otu_df.columns]
y = data["group"].map({"Control": 0, "Treatment": 1})
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Train SVM
model = SVC(kernel='linear', probability=True, random_state=42)
model.fit(X_train, y_train)

# Predict and evaluate
y_pred = model.predict(X_test)
acc = accuracy_score(y_test, y_pred)
cm = confusion_matrix(y_test, y_pred)

print(f"Accuracy: {acc:.2f}")
print("Confusion Matrix:")
print(cm)
```

## R Code (caret)
```{r}
library(tidyverse)
library(caret)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")
otu_df <- otu_df[, meta_df$sample_id]
otu_df <- t(otu_df)
data <- cbind(as.data.frame(otu_df), group = meta_df$group)

# Encode group and split
data$group <- as.factor(data$group)
set.seed(42)
trainIndex <- createDataPartition(data$group, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]

# Train SVM
svm_model <- train(group ~ ., data = train, method = "svmLinear", trControl = trainControl(method = "cv", number = 5))

# Predict and evaluate
pred <- predict(svm_model, newdata = test)
confusionMatrix(pred, test$group)
```

# How do I apply Gradient Boosting (XGBoost) for microbiome classification?

## Explanation

**XGBoost** is an efficient, scalable gradient boosting algorithm widely used in biological classification tasks. It's known for:
- High performance on structured/tabular data
- Handling of missing values
- Built-in feature importance metrics

This Q&A shows how to train and evaluate an XGBoost classifier.

## Python Code

```python
import pandas as pd
from xgboost import XGBClassifier
from sklearn.metrics import accuracy_score, confusion_matrix
from sklearn.model_selection import train_test_split

# Load and prepare data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")
X = data[otu_df.columns]
y = data["group"].map({"Control": 0, "Treatment": 1})
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Train XGBoost classifier
model = XGBClassifier(use_label_encoder=False, eval_metric='logloss', random_state=42)
model.fit(X_train, y_train)

# Predict and evaluate
y_pred = model.predict(X_test)
acc = accuracy_score(y_test, y_pred)
cm = confusion_matrix(y_test, y_pred)

print(f"Accuracy: {acc:.2f}")
print("Confusion Matrix:")
print(cm)

# Top important OTUs
importances = pd.Series(model.feature_importances_, index=X.columns)
print("Top Features:
", importances.sort_values(ascending=False).head())
```

## R Code (caret + xgboost)
```{r}
library(tidyverse)
library(caret)
library(xgboost)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")
otu_df <- otu_df[, meta_df$sample_id]
otu_df <- t(otu_df)
data <- cbind(as.data.frame(otu_df), group = meta_df$group)

# Encode group and split
data$group <- as.factor(data$group)
set.seed(42)
trainIndex <- createDataPartition(data$group, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]

# Train XGBoost
xgb_model <- train(group ~ ., data = train, method = "xgbTree", trControl = trainControl(method = "cv", number = 5))

# Predict and evaluate
pred <- predict(xgb_model, newdata = test)
confusionMatrix(pred, test$group)
```

# How do I visualize ROC curves to compare classification models?

## Explanation

**Receiver Operating Characteristic (ROC) curves** help visualize model performance across different thresholds. The **Area Under the Curve (AUC)** summarizes performance — closer to 1.0 is better.

Comparing ROC curves across models (e.g., Random Forest, Logistic Regression, XGBoost) provides insight into which performs best and where they differ.

This Q&A demonstrates ROC curve generation in Python and R.

## Python Code
```python
import pandas as pd
from sklearn.model_selection import train_test_split
from sklearn.ensemble import RandomForestClassifier
from sklearn.linear_model import LogisticRegression
from xgboost import XGBClassifier
from sklearn.metrics import roc_curve, auc
import matplotlib.pyplot as plt

# Load and prepare data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")
X = data[otu_df.columns]
y = data["group"].map({"Control": 0, "Treatment": 1})
X_train, X_test, y_train, y_test = train_test_split(X, y, test_size=0.3, random_state=42)

# Define models
models = {
    "Random Forest": RandomForestClassifier(random_state=42),
    "Logistic Regression": LogisticRegression(max_iter=1000),
    "XGBoost": XGBClassifier(use_label_encoder=False, eval_metric="logloss", random_state=42)
}

# Plot ROC curves
plt.figure(figsize=(8, 6))
for name, model in models.items():
    model.fit(X_train, y_train)
    probas = model.predict_proba(X_test)[:, 1]
    fpr, tpr, _ = roc_curve(y_test, probas)
    auc_score = auc(fpr, tpr)
    plt.plot(fpr, tpr, label=f"{name} (AUC = {auc_score:.2f})")

plt.plot([0, 1], [0, 1], "k--")
plt.xlabel("False Positive Rate")
plt.ylabel("True Positive Rate")
plt.title("ROC Curve Comparison")
plt.legend()
plt.tight_layout()
plt.show()
```

## R Code (caret + pROC)
```{r eval=FALSE}
library(tidyverse)
library(caret)
library(pROC)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")
otu_df <- otu_df[, meta_df$sample_id]
otu_df <- t(otu_df)
data <- cbind(as.data.frame(otu_df), group = as.factor(meta_df$group))

# Train/test split
set.seed(42)
trainIndex <- createDataPartition(data$group, p = .7, list = FALSE)
train <- data[trainIndex, ]
test  <- data[-trainIndex, ]

# Define fixed tuning for models that require it
rf_grid <- data.frame(mtry = floor(sqrt(ncol(train) - 1)))
svm_grid <- data.frame(C = 1)
xgb_grid <- data.frame(
  nrounds = 50,
  max_depth = 3,
  eta = 0.3,
  gamma = 0,
  colsample_bytree = 1,
  min_child_weight = 1,
  subsample = 1
)

# Train models using fixed tuneGrid
models <- list(
  rf = train(group ~ ., data = train, method = "rf", trControl = trainControl(method = "none"), tuneGrid = rf_grid),
  glm = train(group ~ ., data = train, method = "glm", family = "binomial", trControl = trainControl(method = "none")),
  svm = train(group ~ ., data = train, method = "svmLinear", trControl = trainControl(method = "none"), tuneGrid = svm_grid),
  xgb = train(group ~ ., data = train, method = "xgbTree", trControl = trainControl(method = "none"), tuneGrid = xgb_grid)
)

# ROC analysis
roc_list <- lapply(models, function(model) {
  probs <- predict(model, newdata = test, type = "prob")
  # Safely extract numeric probabilities for the "Treatment" class
  class_label <- "Treatment"
  if (!(class_label %in% colnames(probs))) {
    stop(paste("Class label", class_label, "not found in predicted probabilities"))
  }
  prob_values <- as.numeric(probs[, class_label])
  roc(response = test$group, predictor = prob_values)
})

# Plot ROC
plot(roc_list[[1]], col = "blue", legacy.axes = TRUE, main = "ROC Curves - Microbiome Classification")
cols <- c("blue", "green", "red", "purple")
for (i in 2:length(roc_list)) {
  plot(roc_list[[i]], col = cols[i], add = TRUE)
}
legend("bottomright", legend = names(models), col = cols, lwd = 2)
```


# How do I apply cross-validation strategies to evaluate model reliability?

## Explanation

**Cross-validation (CV)** is critical for evaluating how well a machine learning model generalizes to unseen data. It reduces the risk of overfitting by testing the model on multiple train/test splits.

Common CV strategies:
- **k-Fold**: Split into k subsets, rotate test set
- **Repeated k-Fold**: More robust by repeating k-fold several times
- **Stratified k-Fold**: Ensures balanced class distribution in folds

This Q&A shows how to apply CV in Python and R using common microbiome classifiers.

## Python Code
```python
import pandas as pd
from sklearn.ensemble import RandomForestClassifier
from sklearn.model_selection import cross_val_score, StratifiedKFold

# Load and prepare data
otu_df = pd.read_csv("data/otu_table_filtered.tsv", sep="\t", index_col=0).T
meta_df = pd.read_csv("data/sample_metadata.tsv", sep="\t")
data = pd.merge(otu_df, meta_df, left_index=True, right_on="sample_id")
X = data[otu_df.columns]
y = data["group"].map({"Control": 0, "Treatment": 1})

# Define CV strategy
cv = StratifiedKFold(n_splits=5, shuffle=True, random_state=42)

# Evaluate Random Forest using cross-validation
model = RandomForestClassifier(random_state=42)
scores = cross_val_score(model, X, y, cv=cv, scoring="accuracy")

print("Cross-Validation Accuracy Scores:", scores)
print("Mean Accuracy:", scores.mean())
```

## R Code (caret with repeated k-fold CV)
```{r}
library(tidyverse)
library(caret)

otu_df <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta_df <- read.delim("data/sample_metadata.tsv")
otu_df <- otu_df[, meta_df$sample_id]
otu_df <- t(otu_df)
data <- cbind(as.data.frame(otu_df), group = as.factor(meta_df$group))

# Define CV control
ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 3)

# Train with CV
set.seed(42)
cv_model <- train(group ~ ., data = data, method = "rf", trControl = ctrl)

# Results
print(cv_model)
plot(cv_model)
```

# How do I use `mikropml` in R for microbiome machine learning?

## Explanation

[`mikropml`](https://github.com/SchlossLab/mikropml) is a microbiome-focused R package by Pat Schloss designed for:
- End-to-end modeling workflows
- Built-in cross-validation and hyperparameter tuning
- Transparency in model reporting and evaluation

It simplifies the process of building, tuning, and interpreting microbiome ML models.

This Q&A introduces a basic pipeline using `mikropml` and prepares the OTU and metadata files as expected.

## R Code

```{r}
# 📦 Ensure mikropml is installed
if (!requireNamespace("mikropml", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
  remotes::install_github("SchlossLab/mikropml")
}

library(mikropml)
library(tidyverse)

# Load OTU table and metadata
otu <- read.delim("data/otu_table_filtered.tsv", row.names = 1)
meta <- read.delim("data/sample_metadata.tsv")

# Transpose OTU so samples are rows
otu_t <- t(otu)
otu_df <- as.data.frame(otu_t)
otu_df$sample_id <- rownames(otu_t)

# Merge with metadata
data <- inner_join(otu_df, meta, by = "sample_id")

# Run mikropml using run_ml()
set.seed(42)
fit <- run_ml(
  dataset = data,
  outcome_colname = "group",
  method = "rf",        # Choose from rf, svm, glmnet, xgb
  seed = 42
)

# View model summary
summary(fit)

# Plot variable importance
fit$importance_plot
```

## Notes
- mikropl supports additional tuning and export for reproducibility.
- `mikropml()` auto-detects classification vs regression tasks.
